---
title: "in-class-for-loop"
output: html_document
date: '2023-10-23'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# QUESTION 1: What is the difference between these two codes?

Option 1 seems more intuitive, but option 2 makes it easier to access the looping index. 

```{r iterate-over-vector-elements}
plants <- c("monstera", "sanseveria", "photos")
for (i in plants) {
  print(i)
}
```

```{r iterate-over-vector-indices}
plants <- c("monstera", "sanseveria", "photos")
for (i in 1:length(plants)) {  # for (i in seq_along(plants))
  print(i)
  print(plants[i])
  print(paste(plants[i], "is on position", i))
}
```


# QUESTION 2: How do you loop over a list instead than a vector?

It is the same but to access elements of lists you need double `[[]]`

```{r iterate-over-list-elements}
plants <- c("monstera", "sanseveria", "photos")
for (i in plants) {
  print(i)
}
```

```{r iterate-over-list-indices}
plants <- list("monstera", "sanseveria", "photos")
for (i in 1:length(plants)) {  # for (i in seq_along(plants))
  print(i)
  print(plants[[i]])
  print(paste(plants[[i]], "is on position", i))
}
```


# QUESTION 3: When to use a list VS a vector? 

Main reasons are different data types, named elements, can change their length
Exercise: modify the code above so using a list makes more sense than using a vector.


# QUESTION 4: How do you loop over columns of a dataframe? 

First, modify the previous code to create a toy dataframe
```{r create-df}
name <- c("monstera", "sanseveria", "photos")
age <- c(5, 2, 1) 
light <- c("moderate", "low", "moderate")
(plants_df <- data.frame(Name = name, Age = age, Light = light)) 
#class(plants_df)
```

Then iterate over its elements...
```{r iterate-over-df-elements}
for (i in plants_df) {
  print(i)
}
```

Does not quite work as expected... why? Instead, iterate over indices which corresponds to columns
```{r}
for (i in 1:length(plants_df)) {  # for (i in seq_along(plants_df))
  print(i)
  #print(plants_df[[i]])
  print(paste(plants_df[[i]], "is in column", i))
}

```


# EXAMPLE OF USING FOR LOOP TO CALCULATE MEAN OF COLUMNS OF PALMERPENGUINS DATAFRAME

Goal: use a for loop to calculate the mean value for each column of a subset of the `palmerpenguins`

Load libraries and data
```{r}
library(tidyverse)
library(palmerpenguins)
data(penguins)
```

Select only variables of interest and remove missing data
```{r}
head(penguins)
penguins_clean <- select(penguins, 3:6) %>% drop_na()
head(penguins_clean)
```

We can reach our goal in two steps: first we calculate the mean value for each column in the data (Step 1), then we save the results (Step 2)

Step 1: calculate the means

Our first instinct might be to write the following code, which seems to work...
```{r}
for (i in penguins_clean) {
  print(mean(i))
}
```

...but it does not work as we might think it does!
We expect the element `i` to be a number that runs from 1 to 4 because there are four columns in our data, but it is not. Let's verify it by printing `i` at every loop iteration
```{r}
for (i in penguins_clean) {
  print(i)
  print(mean(i))
}
```

Instead, write code such that at every iteration of the loop the element `i` varies and is a different column
```{r}
for (i in 1:length(penguins_clean)) {  # same as 1:4
  print(i)
  print(mean(penguins_clean[[i]]))
}
```

Notice the use of`[[]]` because columns are lists: we need to access their elements
```{r}
is.list(penguins_clean)
```

Step 2: save the means we calculated for each column

To begin with, use the `<-` operator to assign results to an object.
We might be tempted to write the following code, but it won't work the way we expect it to work. Why?
```{r}
for (i in 1:length(penguins_clean)) {  
  output <- (mean(penguins_clean[[i]]))
}
output
```

Instead, first initialize an output vector outside the loop (notice `mode = "double"`: to initialize other types replace with "logical", "character" etc.)
```{r}
output <- vector(mode = "double" , length = length(penguins_clean)) 
output
```

Then, loop through each column of the data frame, and each time we save the result in the output vector (notice here you do not need the `[[]]` for output, `[]` is enough, but they are often added for code syntax consistency)
```{r}
for (i in 1:length(penguins_clean)) {
  output[[i]] <- mean(penguins_clean[[i]])
  print(output[[i]])
}
output
```

The code below shows the same operation using `seq_along()` which is a better option than `1:length(object)`: it creates a vector that contains a sequence of numbers from 1 to the length of the object,
```{r}
for (i in seq_along(penguins_clean)) {
  output[[i]] <- mean(penguins_clean[[i]])
}
output
```
