---
title: "demo-for-loop"
output: html_document
date: '2023-10-23'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Example of using a for loop with a data frame. Goal: use a for loop to calculate the mean value for each column of a subset of the `palmerpenguins` data


Load libraries and data
```{r}
library(tidyverse)
library(palmerpenguins)
data(penguins)
```


Perform some initial data cleaning operations: select only variables of interest and remove missing data
```{r}
head(penguins)
penguins_clean <- select(penguins, 3:6) %>% drop_na()
head(penguins_clean)
```


We can reach our goal in two steps: first we calculate the mean value for each column in the data (task 1), then we save the results (task 2)


TASK 1: calculate the means

Our first instinct might be to write the following code, which seems to work...
```{r}

```


...but the syntax is actually incorrect: the element `i` should be a number that runs from 1 to 4 (there are four columns in our data) but it is not, and we can verify it by printing `i` at every loop iteration
```{r}

```


We should write code such that at every iteration of the loop the element `i` varies and it is a different column, like that:
```{r}

```

Notice we use the `[[]]` because columns are lists: we need to access their elements
```{r}

```


TASK 2: save the means

To begin with, we want to use the `<-` operator to assign results to an object.

Our first instinct might be to write the following code, but it won't work, why?
```{r}

```


Instead, we first initialize an output vector outside the loop (notice `mode = "double"`: to initialize other types replace with "logical", "character" etc.)
```{r}

```


Then, we loop through each column of the data frame, and each time we save the result in the output vector (notice here you do not need the `[[]]` for output, `[]` is enough, but they are often added for code syntax consistency)

```{r}

```
And we are done!


The code below shows the same operation using `seq_along()` which is a better option than `1:length(object)`: it creates a vector that contains a sequence of numbers from 1 to the length of the object,
```{r}

```


